% File: mas_report.tex
\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{lmodern}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{microtype}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
\makeatletter
\renewenvironment{thebibliography}[1]
     {\section*{References\@mkboth{REFERENCES}{REFERENCES}}%
      \list{\@biblabel{\arabic{enumiv}}}%
           {\settowidth\labelwidth{\@biblabel{#1}}%
            \leftmargin\labelwidth
            \advance\leftmargin\labelsep
            \usecounter{enumiv}%
            \itemsep .01in}%
      \sloppy\clubpenalty4000\widowpenalty4000
      \sfcode`\.=1000\relax}
     {\endlist}
\makeatother
\pdfinfo{
/Title (TileWorld Multi-Agent System)
/Author (Ananya Singh, Twissa Modi, Aditya Kumar Goel, Yu Taek Lee, Nik Shahrul Nafiz Bin Nik Azhan, Mate Tringer)}
\setcounter{secnumdepth}{0}
\setlength\titlebox{3.5in}

\begin{document}

\title{TileWorld Multi-Agent System}
\author{
AI6125 Multi-Agent Systems \quad Group 18\\[4pt]
Ananya Singh \quad Twissa Modi \quad Aditya Kumar Goel\\
Yu Taek Lee \quad Nik Shahrul Nafiz Bin Nik Azhan \quad M\'ate Tringer
}
\maketitle

\begin{abstract}
\begin{quote}
We present a six-agent multi-agent system built for the TileWorld simulation
using the MASON framework. TileWorld is a challenging benchmark in which agents
must explore a partially observable grid, discover a hidden fuel station, collect
tiles, and fill holes to earn reward --- all under tight fuel constraints and with
objects that expire after a limited lifetime. Our approach uses a two-layer
architecture: a shared infrastructure layer that provides A* pathfinding,
fuel-aware planning, memory with observation timestamps, and a broadcast
communication protocol, plus a specialisation layer of six role-specific agents
built on top of it. Through four development phases, average reward on Config~1
improved from 32.1 to 864.3. A 10-run Phase~4 validation on Config~2 achieved an average reward of 888.6
with zero failures, though Config~2 carries structural reliability risk not
fully captured by a small sample. The largest single gain came from introducing multi-agent
coordination in Phase~3, while Phase~4 added role-specific behaviour grounded in
the bold-vs-cautious agent framework of Pollack and Ringuette (1990).
\end{quote}
\end{abstract}

% -----------------------------------------------------------------------
\section{Introduction}

TileWorld is a well-known testbed for evaluating agent architectures under
partial observability and resource constraints \cite{pollack1990}. Six agents
share a bounded grid and must cooperate to maximise the number of tiles
successfully delivered into holes within a fixed simulation budget.
The environment presents several concurrent challenges: agents can only see a
small local area, so global map knowledge must be built incrementally through
movement and communication; fuel is consumed on every move and can only be
replenished at a single hidden station; and both tiles and holes expire after a
limited number of steps, making timing and freshness critical.

A naive solution --- six independent agents each running the same generic loop
--- wastes effort through redundant exploration, duplicate targeting, and
uninformed routing. Our goal was to design a system that behaves as a
coordinated team while keeping each agent's design tractable and individually
testable.

We address this with a \emph{two-layer architecture}. The first layer is shared
infrastructure implemented in three classes: \texttt{SmartTWAgent} (the base
agent), \texttt{SmartTWAgentMemory} (time-stamped shared memory), and
\texttt{SmartTWPlanner} (goal-directed, fuel-aware planning). Every improvement
to this layer benefits all six agents simultaneously. The second layer is a set
of specialist agents that extend the base class and each prioritise a different
aspect of the task: fuel discovery, tile collection, delivery reliability, map
coverage, high-value routing, and failure prevention.

The rest of this paper is organised as follows. Section~2 places the work in the
context of TileWorld research. Section~3 describes the shared architecture.
Section~4 covers the communication protocol. Section~5 presents each specialist
agent. Section~6 describes the experimental configurations. Section~7 reports
results. Section~8 analyses strengths and weaknesses, and Section~9 concludes.

% -----------------------------------------------------------------------
\section{Background}

TileWorld was introduced by Pollack and Ringuette \shortcite{pollack1990} as an
experimental platform for studying agent deliberation. Their central concern was
the trade-off between \emph{bold} agents that commit to plans and execute them
without interruption, and \emph{cautious} agents that reconsider their plan at
every step. They found that neither extreme dominates: bold agents waste effort
reconsidering, while cautious agents miss opportunities by abandoning plans too
eagerly. The optimal balance depends on the environment's rate of change.

Subsequent work surveyed in \cite{kinny1993} confirmed that the right level of
commitment varies with environmental volatility. High-volatility settings (many
objects, short lifetimes) reward cautious behaviour; low-volatility settings
(sparse, long-lived objects) reward bold commitment.

Our architecture operationalises this insight directly. The shared planner uses
different affordability margins depending on whether the runtime environment is
detected as short-lifetime or stable. The \texttt{SmarterReplanningAgent}
specialist functions as a configurable cautious agent: in volatile conditions it
projects whether a target will survive until arrival and abandons the plan
proactively if not; in stable conditions it uses a looser threshold and commits
longer.

% -----------------------------------------------------------------------
\section{System Architecture}

\subsection{Two-Layer Design}

The architecture separates concerns into two layers.

\textbf{Layer~1 --- Shared Foundation.}
All six agents inherit from \texttt{SmartTWAgent} and use the same
\texttt{SmartTWAgentMemory} and \texttt{SmartTWPlanner}. This layer provides:
A* pathfinding with obstacle avoidance and fuel awareness;
a memory store with per-cell observation timestamps and automatic staleness
tracking;
a claim system to prevent two agents targeting the same object;
runtime environment classification (\texttt{isDense()},
\texttt{isShortLifetime()}, \texttt{isLargeGrid()});
and a full broadcast communication stack.

\textbf{Layer~2 --- Specialist Agents.}
Each specialist extends \texttt{SmartTWAgent} and overrides the \texttt{think()}
method to insert role-specific decisions before delegating to the shared base
logic via \texttt{super.think()}. This keeps each specialist focused on its role
while retaining all base capabilities unchanged.

\subsection{Shared Agent Loop}

Every simulation step follows a fixed ordering enforced by the MASON scheduler:

\begin{enumerate}
\item \textbf{Environment step} (ordering 1): new objects are created, expired
objects are removed, and the message list is cleared.
\item \textbf{Sense and communicate} (ordering 2): each agent observes nearby
cells, updates memory, reads incoming messages, and broadcasts its own.
\item \textbf{Think and act} (ordering 3): each agent runs its decision logic
and executes one action (move, pickup, drop, refuel, or wait).
\end{enumerate}

Within ordering~2, agents run \texttt{sense(); communicate();} sequentially in
creation order, each reading the message queue before appending to it. A message
broadcast by agent~$N$ is therefore visible during the same step to agents
$N{+}1$ through~6 (which read after $N$ writes), but not to agents~$1$
through~$N{-}1$ (which already read before $N$ wrote). By ordering~3, all
broadcasts from ordering~2 are in the queue; however, each agent's stored
parsed data reflects only the messages that existed when it called
\texttt{communicate()}, so earlier-ordered agents see later-ordered agents'
broadcasts at the following step rather than the current one. In practice this means fuel station discoveries
and tile or hole sightings propagate to the full team by the following step at
the latest, and to later-ordered agents within the same step.

\subsection{A* Pathfinding}

Navigation is handled by \texttt{AstarPathGenerator}, which computes
obstacle-avoiding paths using the standard A* algorithm with Manhattan distance
as the heuristic. The planner wraps path generation with an affordability check:
before committing to a goal, it verifies that the agent has enough fuel to reach
the target and return to the fuel station. In short-lifetime environments the
planner also rejects targets whose estimated remaining lifetime is shorter than
the steps required to arrive.

\subsection{Memory and Environment Detection}

\texttt{SmartTWAgentMemory} stores observation timestamps for every seen cell,
tile, and hole. These timestamps drive three runtime signals:

\begin{itemize}
\item \texttt{isShortLifetime()} --- true when observed object disappearance
rates suggest objects expire in roughly 30 steps or fewer.
\item \texttt{isDense()} --- true when the observed object creation rate is high.
\item \texttt{isLargeGrid()} --- true when the map appears to exceed roughly
60 cells on a side.
\end{itemize}

These signals are estimated from observations rather than hardcoded to a
specific configuration, so the same agents adapt correctly to both Config~1 and
Config~2 without any parameter changes.

% -----------------------------------------------------------------------
\section{Communication Protocol}

Agents communicate via a global broadcast channel. Because ordering~2 runs
agents sequentially, a broadcast from agent~$N$ reaches later-ordered agents
within the same step and earlier-ordered agents by the following step at the
latest; the full team has the message within one step. The base class parses all
incoming messages and stores results in shared protected fields, so specialist
agents consume parsed data directly without writing any parsing code themselves.

Table~\ref{tab:messages} lists the nine message types.

\begin{table*}[t]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Type} & \textbf{Format} & \textbf{Purpose} \\
\midrule
\texttt{FUEL}     & \texttt{x,y}           & Fuel station location \\
\texttt{TILE}     & \texttt{x,y,time}      & New tile sighting \\
\texttt{HOLE}     & \texttt{x,y,time}      & New hole sighting \\
\texttt{GONE}     & \texttt{x,y}           & Object disappeared \\
\texttt{CLAIM}    & \texttt{x,y}           & Agent is targeting this cell \\
\texttt{LOW}      & \texttt{agentId,fuel}  & Low-fuel warning \\
\texttt{EXPIRING} & \texttt{x,y,reason}    & Target likely to expire \\
\texttt{HOTSPOT}  & \texttt{x,y,density}   & High-density area \\
\texttt{SWAP}     & \texttt{agentId,status}& Zone exploration complete \\
\bottomrule
\end{tabular}
\caption{The nine message types and their payloads.}
\label{tab:messages}
\end{table*}

Communication improves team performance in four ways. First, \textbf{faster
discovery sharing}: fuel station discovery by any one agent immediately benefits
all five others. Second, \textbf{deduplication}: CLAIM messages prevent two
agents from wasting movement on the same target. Third, \textbf{freshness
control}: EXPIRING warnings let all agents avoid routing toward targets that are
about to vanish. Fourth, \textbf{coverage coordination}: HOTSPOT and SWAP
messages guide exploration toward productive and uncovered regions.

% -----------------------------------------------------------------------
\section{Specialist Agents}

Each specialist inherits all base capabilities and adds a focused decision layer
in \texttt{think()}.

\subsection{FuelScoutAgent}

\texttt{FuelScoutAgent} is designed to accelerate fuel station discovery.
While the station is unknown, it uses a search radius of 20 cells ---
significantly wider than the base agent's lawnmower pattern --- to jump between
zones and cover the map faster. Its refuel threshold adapts to grid size
(\texttt{isLargeGrid() ? 250 : 175}) to ensure it never runs critically low
before the station is found. Once fuel is known and broadcast, it reverts to
standard base behaviour. It also sends \texttt{LOW} warnings when fuel drops
below 150, alerting teammates to shift coverage away from fuel-scarce regions.

\subsection{TileHunterAgent}

\texttt{TileHunterAgent} focuses on efficient tile collection. Its key behaviour
is adaptive batching: in stable, long-lifetime environments it batches up to
three tiles before delivering, amortising the fuel cost of a delivery trip over
three reward points; in short-lifetime environments it delivers after every
single tile to avoid objects expiring in transit. It immediately claims any tile
it targets to reduce collisions and uses \texttt{HOTSPOT} data to bias movement
toward tile-dense regions.

\subsection{HoleFillerAgent}

\texttt{HoleFillerAgent} protects the team from wasted delivery trips. When
carrying a tile it checks whether the nearest hole has been flagged as expiring
by a teammate; if so, it voids the current plan and seeks an alternative.
In dense environments it patrols within five cells of a known hole while empty,
reducing the distance it must travel once it acquires a tile. In sparse
environments it skips patrol entirely, since the fuel cost of circling outweighs
the benefit. It also rebroadcasts fresh hole locations at intervals to keep the
team's delivery knowledge current.

\subsection{ExplorerAgent}

\texttt{ExplorerAgent} ensures systematic map coverage. It uses
\texttt{getLeastVisitedTarget()} with a search radius that adapts to grid size
(\texttt{isLargeGrid() ? 20 : 10}) to always move toward the least-recently-seen
area in its assigned zone. It biases movement toward \texttt{HOTSPOT} positions
broadcast by the \texttt{DeliveryOptimizerAgent}, but only when the hotspot lies
within its own zone. When coverage exceeds 80\% it broadcasts a \texttt{SWAP}
message to signal zone completion, allowing the team to shift exploration
capacity elsewhere.

\subsection{DeliveryOptimizerAgent}

\texttt{DeliveryOptimizerAgent} maximises the reward per fuel unit. When
carrying a tile, it selects the target hole by scoring all known holes on three
factors: distance from current position, freshness (age relative to lifetime),
and proximity to \texttt{HOTSPOT} regions where additional holes and tiles are
likely to appear. This multi-gate scoring avoids the pathological cases of pure
nearest-hole routing (choosing a hole that is about to expire) and pure
freshness routing (choosing a distant hole over a nearby one). It also
broadcasts \texttt{HOTSPOT} messages from high-density areas it observes,
sharing information that directly improves the Explorer's coverage decisions.

\subsection{SmarterReplanningAgent}

\texttt{SmarterReplanningAgent} implements the adaptive cautious agent concept
from Pollack and Ringuette \shortcite{pollack1990}. At each step it applies
three checks on any active plan. First, it checks whether a teammate has already
flagged the current goal as expiring; if so, it voids the plan immediately.
Second, it independently projects whether the goal will still exist on arrival
using observation timestamps and estimated remaining lifetime; the expiry
threshold adapts to environment volatility (\texttt{isShortLifetime() ? 0.7 :
0.9}). Third, it checks whether the agent has enough fuel to reach the goal and
return to the fuel station, with a safety margin that scales with grid size
(\texttt{max(50, gridWidth / 4)}). Failed checks trigger early replanning rather
than waiting for the plan to fail at execution time, and the agent broadcasts an
\texttt{EXPIRING} message to warn teammates of bad targets it detects.

% -----------------------------------------------------------------------
\section{Experimental Configurations}

The system was evaluated on two configurations derived from the MASON
TileWorld parameters. Table~\ref{tab:configs} summarises their key differences.

\begin{table*}[t]
\centering
\small
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Parameter} & \textbf{Config~1} & \textbf{Config~2} \\
\midrule
Grid size       & 50 $\times$ 50 & 80 $\times$ 80 \\
Tile rate (per step)     & 0.2  & 2.0 \\
Hole rate (per step)     & 0.2  & 2.0 \\
Obstacle rate (per step) & 0.2  & 2.0 \\
Object lifetime (steps)  & 100  & 30  \\
Agent fuel               & 500  & 500 \\
\bottomrule
\end{tabular}
\caption{Parameters for Config~1 and Config~2.}
\label{tab:configs}
\end{table*}

Config~1 is the easier environment: objects are sparse but long-lived, so
careful planning and commitment pay off. Config~2 is harder: the larger grid
puts more pressure on fuel, and the short object lifetime means memory
information becomes stale quickly, increasing the cost of a bad plan.

% -----------------------------------------------------------------------
\section{Results}

\subsection{Phase Development (Config~1)}

Table~\ref{tab:phases} shows the average reward achieved at each development
phase, measured on Config~1. Each phase added a distinct capability.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrrp{3.2cm}@{}}
\toprule
\textbf{Phase} & \textbf{Avg} & \textbf{Min} & \textbf{Max} & \textbf{Key addition} \\
\midrule
Phase~1 & 32.1  & 0   & 55  & Survival and memory \\
Phase~2 & 160.3 & 0   & 249 & A* planning, delivery \\
Phase~3 & 541.5 & 469 & 601 & Coordination, communication \\
Phase~4 & 864.3 & 801 & 959 & Adaptive base + role specialisation \\
\bottomrule
\end{tabular}
\caption{Config~1 average reward by development phase. Phases~1--3 are frozen
historical measurements from earlier codebase states; Phase~4 is a current
rerun on the active codebase after all follow-up commits.}
\label{tab:phases}
\end{table}

Phase~1 established a survival baseline: \texttt{SmartTWAgent} and
\texttt{SmartTWAgentMemory} replaced the default agent with fuel-aware movement,
memory decay, and opportunistic tile pickup. The system reacted to local events
but could not yet route toward high-value work.

Phase~2 added \texttt{SmartTWPlanner} with A* pathfinding. Agents stopped
wandering and began executing goal-directed plans. The first major reward jump
followed because the same map knowledge was now exploited efficiently.

Phase~3 introduced multi-agent coordination: zone-based exploration,
message broadcasts, shared tile and hole knowledge, and claim-based
deconfliction. The score jump from Phase~2 to Phase~3 (160.3 to 541.5, a 3.4$\times$
increase) was the largest in the project and came from structural rather than
incremental improvement: six agents began working as a team rather than as
independent individuals.

Phase~4 unfolded in three steps. First, adaptive improvements to
\texttt{SmartTWAgent} and \texttt{SmartTWAgentMemory} directly: runtime
environment detection, adaptive fuel safety margins, and smarter planning
heuristics. Second, a two-stage architectural transition: the uniform agent loop was first
broken into individually instantiated person-named subclasses (one per team
member), then renamed to role-named classes (\texttt{TileHunterAgent},
\texttt{HoleFillerAgent}, etc.), both stages initially delegating entirely to
the shared base. Third, follow-up commits
added the substantive role-specific logic --- tighter expiry thresholds,
freshness-aware delivery scoring, reachability checks, and throttled broadcasts.
The 864.3 average reflects this full three-step progression; the largest gains
came from the adaptive base improvements and the coordination already established
in Phase~3, with role-specific behaviour providing targeted quality refinements
on top.

\subsection{Final Validation (Phase~4)}

Table~\ref{tab:results} shows the Phase~4 validation results on both
configurations.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Config} & \textbf{Avg reward} & \textbf{Final reward} & \textbf{Failures} \\
\midrule
Config~1 & 864.3 & 959 & 0 \\
Config~2 & 888.6 & 981 & 0 \\
\bottomrule
\end{tabular}
\caption{Phase~4 validation results. Each config was evaluated with
\texttt{TileworldMain} running 10 independent random-seed trials
(\texttt{iteration = 10}; see \texttt{TileworldMain.java:30}). Average reward
is the mean final score across the 10 trials; Final reward is the score of the
last trial. A failure is a trial whose final reward was zero. All three
aggregate metrics are recorded in \texttt{MULTI\_RUN\_RESULTS.md}. Per-trial
variance (individual min/max across the 10 trials) was not separately recorded
for Config~2.}
\label{tab:results}
\end{table}

Both 10-run validations completed with zero failures. Config~2 returned a
slightly higher average (888.6 vs.\ 864.3), but Config~1 remains the clearest
success case because its environment is more amenable to the coordination and
planning strategies the system relies on. Config~2 structural risks are
discussed in the analysis below.

% -----------------------------------------------------------------------
\section{Analysis}

\subsection{Strengths}

\textbf{Shared-base scalability.} Implementing pathfinding, memory, and
communication once in the base class means every algorithmic improvement
benefits all six agents simultaneously. This was the most important engineering
decision in the project.

\textbf{Coordination under partial observability.} Without communication, each
agent is limited to what it can directly observe. The broadcast protocol
converts six local observers into a distributed shared-intelligence system.
Claims prevent duplicated effort; EXPIRING warnings stop wasted movement;
HOTSPOT messages direct exploration toward productive regions.

\textbf{Measured performance trajectory.} The phase scores confirm that each
architectural addition produced a measurable gain. The largest gain (Phase~3
coordination) accounts for the majority of total improvement, validating the
design priority of getting coordination right before tuning specialisation.

\textbf{Cross-agent idea borrowing.} Several mechanisms were refined across
agents. For example, \texttt{DeliveryOptimizerAgent} explicitly combines expiry
projection (originally developed for \texttt{SmarterReplanningAgent}), hotspot
routing, and freshness-aware scoring into a single delivery policy. Good ideas
spread through the team incrementally.

\subsection{Weaknesses}

\textbf{Config~2 structural risk.} The 10-run validation produced a positive
average reward (888.6), but the environment's structure creates latent reliability
risk not fully captured by the aggregate. The 80$\times$80 grid creates a larger
search space for the fuel station; delayed discovery forces the whole team into
conservative routing. Short object lifetimes increase the chance that remembered
targets expire before an agent arrives, even with predictive expiry checking. Per-trial variance (individual min/max across the 10 trials) was not separately
recorded for Config~2, so the spread of outcomes is unquantified.

\textbf{Early-run lifetime detection.} The runtime environment classifier
requires enough observed disappearances to estimate object lifetime accurately.
In the first hundred steps of a Config~2 run, agents may still believe they
are in a stable environment and apply overly generous planning horizons.

\textbf{Fuel station discovery bottleneck.} Many decisions across the system
depend on the fuel station being found early. On a large map with an unlucky
starting layout, delayed discovery creates cascading conservative behaviour
across all agents until the station is located.

\subsection{Coordination Tradeoff}

The fundamental tradeoff in any multi-agent system is between the parallelism
benefit --- six agents covering the map simultaneously, collecting tiles, and
filling holes in parallel --- and the coordination cost of communication,
claim management, and zone assignment overhead. This system is strongest when
the environment gives coordination time to pay off: Config~1, with its stable,
sparse objects, rewards careful planning and reliable delivery. Config~2 tests
the limits because the environment changes fast enough that shared information
can become stale before the team acts on it.

% -----------------------------------------------------------------------
\section{Conclusion}

We presented a six-agent TileWorld multi-agent system that improves from a
survival baseline (Phase~1, avg.\ 32.1) to a coordinated specialist team
(Phase~4, avg.\ 864.3 on Config~1) through four development phases.
A 10-run Phase~4 validation on Config~2 achieved an average reward of 888.6;
the four-phase development history is documented for Config~1 only. The key design insight is the two-layer separation between shared
infrastructure and role specialisation: improvements to the base layer compound
across all agents, while specialist layers add targeted decision quality without
duplicating infrastructure.

The most impactful single change was the introduction of multi-agent
coordination in Phase~3, confirming the importance of communication and
deconfliction over individual agent optimisation. Phase~4 role specialisation
added further gains by covering different failure modes of the task: fuel
discovery, tile collection efficiency, delivery reliability, map coverage,
high-value routing, and predictive failure prevention.

Future work could improve Config~2 reliability through two main directions:
earlier fuel station discovery via a dedicated fast-scout strategy that
temporarily overrides zone assignments at the start of a run, and adaptive
lifetime thresholds that update more aggressively in the first 50 steps based
on early disappearance observations.

% -----------------------------------------------------------------------
\bibliographystyle{aaai}
\begin{thebibliography}{}

\bibitem{pollack1990}
Pollack, M.\ E., and Ringuette, M.\ 1990.
Introducing the Tileworld: Experimentally evaluating agent architectures.
In \textit{Proceedings of the Eighth National Conference on Artificial
Intelligence (AAAI-90)}, 183--189.

\bibitem{kinny1993}
Kinny, D., and Georgeff, M.\ 1991.
Commitment and effectiveness of situated agents.
In \textit{Proceedings of the Twelfth International Joint Conference on
Artificial Intelligence (IJCAI-91)}, 82--88.

\end{thebibliography}

\end{document}
